### 2019-04-16 1-Preamble/4-Acknowledgement/acknowledgement.tex
Small changes on Chapter:4 after the talk of yesterday.

I would also like to thank some unusual things. Some people thank God in their thesis, I feel that I prefer to thank the Earth, and apologize for the too many times I took a plane for a conference abroad, and in general about the environment impact of this research work.
I also thank my \href{https://perso.crans.org/besson/zero-dechet/}{bike} which I enjoyed riding everyday despite the too many accidents and rain, my \href{https://perso.crans.org/besson/cuisine/}{enthusiastic daily cooking} who gave me energy (\eg, cheese, chocolate, coffee, vegetables etc), and finally my \href{https://perso.crans.org/besson/}{favorite technological tools}: git, GitHub and Bitbucket, \LaTeX, Bash and Python, GNU/Linux and XUbuntu, Firefox, Konsole and Visual Studio Code, and many other tools.

>>>>

I would also like to thank some unusual things. Some people thank God in their thesis, I feel that I prefer to thank the Earth, and apologize for the too many times I took a plane for a conference abroad, and in general apologize for the environmental impact of this research work. I'll try hard to do better from now on!
I also thank my \href{https://perso.crans.org/besson/zero-dechet/}{bike} which I enjoyed riding everyday despite the too many accidents and rain, my \href{https://perso.crans.org/besson/cuisine/}{enthusiastic daily cooking} who gave me energy (\eg, cheese, chocolate, coffee, vegetables etc), and finally my \href{https://perso.crans.org/besson/}{favorite technological tools}: git, GitHub and Bitbucket, \LaTeX, Bash and Python, GNU/Linux and XUbuntu, Firefox, Konsole and Visual Studio Code, and many other free and open-source tools.

### 2019-04-16 2-Chapters/4-Chapter/IEEE_WCNC_2019__DemoICT.git/main.tex
Small changes on Chapter:4 after the talk of yesterday.

@@

\paragraph{Video}
%
As shown below in Figure~\ref{fig:42:screenshotDemoYouTube},
we realized and released a $6$-minute \textbf{video} to show our demonstration online, and it is available on the YouTube hosting platform, at \texttt{\href{https://youtu.be/HospLNQhcMk}{youtu.be/HospLNQhcMk}}.
The video shows examples of $3$ dynamic devices learning simultaneously, confirming the results of Figure~\ref{fig:42:plot_datafile_append_Uniform_vs_UCB_vs_TS} for overlapping channels.
It also shows the connections between the USRP boards, the Octoclock, the master laptop etc, completing the presentation of the SCEE testbed already shown in Figure~\ref{fig:42:photosSCEETestBed}.

>>>>

\paragraph{Related work.}
% FIXME
\TODOL{Est-ce que je fais ça ici ou à la fin du chapitre 5 ? Ou peut-être en chapitre 1, ça ferait du contenu au chapitre 1...}
We evoke quickly here some previous works that similarly implemented proof-of-concepts to show that reinforcement learning algorithms can be used within real-world wireless communications.
\cite{darak2016bayesian}
\cite{Darak16}
\cite{modiDemo2016}
\cite{kumar2016two}
\cite{KumarYadav2018}
\cite{SawantKumar2018}
\cite{JoshiKumar2018}

@@

\paragraph{Video}
%
As depicted in Figure~\ref{fig:42:screenshotDemoYouTube},
we realized a $6$-minute \textbf{video} to sum-up our demonstration and advertise our work online, and it is available on the YouTube hosting platform, at \texttt{\href{https://youtu.be/HospLNQhcMk}{youtu.be/HospLNQhcMk}}.
The video shows examples of $3$ dynamic devices learning simultaneously, confirming the results of Figure~\ref{fig:42:plot_datafile_append_Uniform_vs_UCB_vs_TS} for overlapping channels.
It also shows the connections between the USRP boards, the Octoclock, the master laptop etc, completing the presentation of the SCEE testbed already shown in Figure~\ref{fig:42:photosSCEETestBed}.
% \textcolor{white}{Of course, the video was not seen much, because our work is useless! Yay!}

### 2019-04-16 2-Chapters/4-Chapter/IEEE_WCNC_2019__DemoICT.git/main_appendix.tex
Small changes on Chapter:4 after the talk of yesterday.

\begin{figure}[!b]
	\centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_TX_PU__v1__simple_grc.png}
    \caption{The \textbf{random traffic generator}. \texttt{Generator} is the only hand-written block, which is configured with a list of active channels, a list of occupancy rate, and constants about the PHY layer (preamble length and data length). It uses one USRP only for emitting data (``Sink'' mode).}
    \label{fig:4app:USRP_TX_PU__v1__simple_grc}
\end{figure}

\begin{figure}[!b]
	\centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_RX_BTS__v1__simple_grc.png}
    \caption{The \textbf{IoT gateway}. The hand-written blocks are the 1) \texttt{Demodulator} block, which is configured with a list of detection threshold (on the received power), 2) the \texttt{Check\_ack} block, which is configured with a maximum block error and an accepted error rates (to decide when a message is close enough to an acknowledgement), and finally 3) the \texttt{send\_ack} block, which is configured with the list of active channels, and knowledge about the PHY layer (delay before sending the Ack and its length). All blocks need to know constants about the PHY layer (preamble length and data length). The IoT gateway uses one USRP for both emitting and receiving data (``Sink'' and ``Source'' modes).}
    \label{fig:4app:USRP_RX_BTS__v1__simple_grc}
\end{figure}

\begin{figure}[!b]
	\centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_TX_SU__v1__simple_grc.png}
    \caption{The \textbf{IoT dynamic object}. The hand-written blocks are the 1) the \texttt{Renormalized\_ack} block, which is configured with the list of active channels and extra knowledge about the PHY layer (preamble length, message length, and a threshold to tune detection of the incoming Ack), 2) the \texttt{Demodulator} and 3) the \texttt{Check\_ack} blocks, both shared with the IoT gateway, and finally 4) the \texttt{generator\_SU} block, which embeds the \UCB{} or Thompson Sampling algorithm, and which is configured with the list of active channels. Most blocks need to know constants about the PHY layer (preamble length and data length). Any of the IoT dynamic objects use one USRP for both emitting and receiving data (``Sink'' and ``Source'' modes).}
    \label{fig:4app:USRP_TX_SU__v1__simple_grc}
\end{figure}

>>>>

For the curious reader, we wanted to include here an illustration and a description of the three components of our demonstration, presented above in Section~\ref{sec:4:gnuradio}.

Figure~\ref{fig:4app:USRP_TX_PU__v1__simple_grc} shows the \textbf{random traffic generator} flow-graph.
\texttt{Generator} is the only hand-written block, which is configured with a list of active channels, a list of occupancy rate, and constants about the PHY layer (preamble length and data length). It uses one USRP only for emitting data (``Sink'' mode).

\begin{figure}[!h]
	\centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_TX_PU__v1__simple_grc.png}
    \caption{The \textbf{random traffic generator} flow-graph.}
    \label{fig:4app:USRP_TX_PU__v1__simple_grc}
\end{figure}

The \textbf{IoT gateway} is presented in Figure~\ref{fig:4app:USRP_TX_PU__v1__simple_grc}.
%
The hand-written blocks are the 1) \texttt{Demodulator} block, which is configured with a list of detection threshold (on the received power), 2) the \texttt{Check\_ack} block, which is configured with a maximum block error and an accepted error rates (to decide when a message is close enough to an acknowledgement), and finally 3) the \texttt{send\_ack} block, which is configured with the list of active channels, and knowledge about the PHY layer (delay before sending the Ack and its length). All blocks need to know constants about the PHY layer (preamble length and data length). The IoT gateway uses one USRP for both emitting and receiving data (``Sink'' and ``Source'' modes).

\begin{figure}[!h]
	\centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_RX_BTS__v1__simple_grc.png}
    \caption{The \textbf{IoT gateway} flow-graph.}
    \label{fig:4app:USRP_RX_BTS__v1__simple_grc}
\end{figure}

Figure~\ref{fig:4app:USRP_TX_SU__v1__simple_grc} below shows the \textbf{IoT dynamic device} flow-graph.
The hand-written blocks are the 1) the \texttt{Renormalized\_ack} block, which is configured with the list of active channels and extra knowledge about the PHY layer (preamble length, message length, and a threshold to tune detection of the incoming Ack), 2) the \texttt{Demodulator} and 3) the \texttt{Check\_ack} blocks, both shared with the IoT gateway, and finally 4) the \texttt{generator\_SU} block, which embeds the \UCB{} or Thompson Sampling algorithm, and which is configured with the list of active channels. Most blocks need to know constants about the PHY layer (preamble length and data length). Any of the IoT dynamic devices use one USRP for both emitting and receiving data (``Sink'' and ``Source'' modes)

\begin{figure}[!h]
	\centering
    \includegraphics[width=1.00\textwidth]{2-Chapters/4-Chapter/Images/USRP_TX_SU__v1__simple_grc.png}
    \caption{The \textbf{IoT dynamic device} flow-graph.}
    \label{fig:4app:USRP_TX_SU__v1__simple_grc}
\end{figure}

### 2019-04-16 2-Chapters/4-Chapter/IEEE_WCNC__2019__Paper__BMBM.git/IEEE_WCNC__2019__Paper__BMBM.tex
Small changes on Chapter:4 after the talk of yesterday.

\TODOL{Remove this part or compress: I will remove as soon as I have clear equations in Chapter~\ref{chapter:2} as references for the \UCB{} algorithm.}

%-----------------------------------------------------------------
\subsection{The \UCB{} algorithm as a building block for different heuristics}
\label{sub:43:MABalgo}
% ----------------------------------------------------------------------

Before presenting our proposed heuristics, we remind here the \UCB{} bandit algorithm \cite{Auer02}.
We include it again to facilitate the understanding of the heuristics, which use \UCB{} as a building block.

% % --- --- --- --- --- --- --- --- --- ---
\paragraph{The \UCB{} algorithm}\label{sub:43:algoUCB}

More formally, for one device, let $N_k(t)$ be the number of times the channel $k$ (for $k\in \llbracket 1, K \rrbracket$) was selected up-to time $t-1$, for $t\geq 0$
for any $t\in\mathbb{N}$,
\begin{equation}\label{eq:43:Nkt}
	N_k(t) = \sum_{\tau=0}^{t-1} \mathbbm{1}(C(\tau) = k),
\end{equation}
The empirical mean estimator $\widehat{\mu_k}(t)$ of channel $k$ is defined as the mean reward obtained up-to time $t-1$,
\begin{equation}\label{eq:43:mukt}
	\widehat{\mu_k}(t) = \frac{1}{N_k(t)} \sum_{\tau=0}^{t-1} r_k(\tau) \mathbbm{1}(C(\tau) = k).
\end{equation}
where $r_{k}(t)$ is the reward obtained after transmission in channel $k$ at time $t$ ($1$ for a successful transmission, and $0$ otherwise)
%
A \emph{confidence} term $B_k(t)$ is given by \cite{Auer02},
\begin{equation}\label{eq:43:Bkt}
	B_k(t) = \sqrt{\alpha \log(t) / N_k(t)},
\end{equation}
where $\alpha$ refers to an exploration coefficient,
% \footnote{~In fact, the larger this coefficient is, the longer the exploration, while the \UCB{} algorithm is proven to be order optimal for $\alpha>0.5$ \cite{Bubeck12}, and has reported a good performance for lower values of $\alpha>0$.},
that we chose equal to $1/2$, as suggested in \cite{Audibert07} and as done in previous works \cite{Bonnefoi18,Bonnefoi17}.
Then, an upper confidence bound in each channel $k$ is defined as
\begin{equation}\label{eq:43:ucb}
	U_k(t) = \widehat{\mu_k}(t) + B_k(t).
\end{equation}
Finally, the transmission channel at time step $t$
is the one maximizing this \UCB{} index $U_k(t)$,
as it is the one expected to be the best one at the current time step $t$,
\begin{equation}\label{eq:43:maxucb}
	C(t) = \arg\max_{1\leq k \leq K} U_k(t).
\end{equation}

The \UCB{} algorithm is implemented independently by each device, and we present it in Algorithm~\ref{algo:43:UCB}.
Note that a device using this first approach is only able to select a channel for the first and all the corresponding retransmissions of a packet.

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			Compute $\forall k, U_k(t) = \widehat{\mu_k}(t) + B_k(t)$, following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
			Transmit in channel $C(t) = \arg\max_{1\leq k \leq K} U_k(t)$\;
			Reward $r_{C(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
		}
		\caption{The \UCB{} algorithm for channel selection (base building block).}
		\label{algo:43:UCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

A device that implements the UCB algorithm is led to focus its transmissions and retransmissions in the channel which is currently identified as the best.
As explained before, focusing in one channel increases the collision probability in retransmissions.
We describe here the proposed heuristics for the channel selection in a retransmission. It is carried out taking
into account that a device can incorporate a different channel selection strategy while being in a back-off state.
Hence, a natural question is to evaluate whether using this additional contextual information can improve the performance of a learning policy.

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^r$}){
				Compute $\forall k, U^r_k(t) = \widehat{\mu^r_k}(t) + B^r_k(t)$ following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
				Transmit in channel $C^r(t) = \arg\max_{1\leq k \leq K} U^r_k(t)$\;
				Reward $r^r_{C^r(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: \UCB{} for retransmission.}    % A naive
		\label{algo:43:TwoUCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		% \KwData{$\forall k,j\in[\![1;K]\!]$, $N_k^j(t)$, $\widehat{\mu_k}^j(t)$, $B_k^j(t)$ and $U_k^j(t)$}
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^j$}){ % Retr. after trying channel $j$
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Compute $\forall k, U_k^j(t) = \widehat{\mu_k}^j(t) + B_k^j(t)$ following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
				Transmit in channel $C^j(t) = \arg\max_{1\leq k \leq K} U^j_k(t)$\;
				Reward $r^j_{C^j(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: $K$ different {\UCB}s for retransmission.}
		\label{algo:43:UCBthenKp1}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

The idea behind this delay is to allow the first stage \UCB{} to start learning the best channel, before starting the second stage \UCB{} (see details in Algorithm~\ref{algo:43:UCBwithDelay}).
The number of transmissions to wait before applying the second algorithm is denoted by $\Delta$, it has to be fixed before-hand\footnote{~Choosing the value of $\Delta$ could be done by extensive benchmarks but such approach goes against the reinforcement learning idea: an heuristic should work against any problem, without the need to be able to simulate the problem in advance in order to find a good value of some internal parameter. As such, we only consider a delay of $\Delta=100$.}.
%
Note that, we use the subscript $({}^d)$ to denote the variables
related to the delayed second-stage \UCB{} algorithm.

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
	\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\uElseIf(\tcp*[f]{Random retransmission}){$t \leq \Delta$}{
				Transmit randomly in a channel $C(t) \sim \mathcal{U}(1,\ldots,K)$.
			}
			\Else(\tcp*[f]{Packet retransmission with delayed $\mathrm{UCB}^d$}){
				Compute $\forall k, U^d_k(t) = \widehat{\mu^d_k}(t) + B^d_k(t)$ following Eqs.~\eqref{eq:43:Nkt},~\eqref{eq:43:mukt}, and~\eqref{eq:43:Bkt}\;
				Transmit in channel $C^d(t) = \arg\max_{1\leq k \leq K} U^d_k(t)$\;
				Reward $r^d_{C^d(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: Delayed \UCB{} for retransmission.}
		\label{algo:43:UCBwithDelay}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

\paragraph{A last idea that we did not explore}
%
Instead of considering a first-stage \UCB{} followed by a random channel retransmission, we could have considered a random first-stage channel retransmission followed by a second-stage \UCB.
This was not considered in our study \cite{Bonnefoi2019WCNC}, and we preferred to not add it to the experiments presented below, for two reasons: to avoid clutter in the plots, and to win some time as these experiments had to run for quite a long time.

>>>>

% %-----------------------------------------------------------------
% \subsection{The \UCB{} algorithm as a building block for different heuristics}
% \label{sub:43:MABalgo}
% % ----------------------------------------------------------------------

% Before presenting our proposed heuristics, we remind here the \UCB{} bandit algorithm \cite{Auer02}.
% We include it again to facilitate the understanding of the heuristics, which use \UCB{} as a building block.

% % % --- --- --- --- --- --- --- --- --- ---
% \paragraph{The \UCB{} algorithm}\label{sub:43:algoUCB}

% More formally, for one device, let $N_k(t)$ be the number of times the channel $k$ (for $k\in \llbracket 1, K \rrbracket$) was selected up-to time $t-1$, for $t\geq 0$
% for any $t\in\mathbb{N}$,
% \begin{equation}\label{eq:43:Nkt}
% 	N_k(t) = \sum_{\tau=0}^{t-1} \mathbbm{1}(C(\tau) = k),
% \end{equation}
% The empirical mean estimator $\widehat{\mu_k}(t)$ of channel $k$ is defined as the mean reward obtained up-to time $t-1$,
% \begin{equation}\label{eq:43:mukt}
% 	\widehat{\mu_k}(t) = \frac{1}{N_k(t)} \sum_{\tau=0}^{t-1} r_k(\tau) \mathbbm{1}(C(\tau) = k).
% \end{equation}
% where $r_{k}(t)$ is the reward obtained after transmission in channel $k$ at time $t$ ($1$ for a successful transmission, and $0$ otherwise)
% %
% A \emph{confidence} term $B_k(t)$ is given by \cite{Auer02},
% \begin{equation}\label{eq:43:Bkt}
% 	B_k(t) = \sqrt{\alpha \log(t) / N_k(t)},
% \end{equation}
% where $\alpha$ refers to an exploration coefficient,
% % \footnote{~In fact, the larger this coefficient is, the longer the exploration, while the \UCB{} algorithm is proven to be order optimal for $\alpha>0.5$ \cite{Bubeck12}, and has reported a good performance for lower values of $\alpha>0$.},
% that we chose equal to $1/2$, as suggested in \cite{Audibert07} and as done in previous works \cite{Bonnefoi18,Bonnefoi17}.
% Then, an upper confidence bound in each channel $k$ is defined as
% \begin{equation}\label{eq:43:ucb}
% 	U_k(t) = \widehat{\mu_k}(t) + B_k(t).
% \end{equation}
% Finally, the transmission channel at time step $t$
% is the one maximizing this \UCB{} index $U_k(t)$,
% as it is the one expected to be the best one at the current time step $t$,
% \begin{equation}\label{eq:43:maxucb}
% 	C(t) = \arg\max_{1\leq k \leq K} U_k(t).
% \end{equation}

% The \UCB{} algorithm is implemented independently by each device, and we present it in Algorithm~\ref{algo:43:UCB}.
% Note that a device using this first approach is only able to select a channel for the first and all the corresponding retransmissions of a packet.

% % \begin{small} % XXX remove if needed
% \begin{figure}[h!]
% 	\centering
%     \begin{framed}
% 	\begin{algorithm}[H]
% 		% \begin{small} % XXX remove if needed
% 		\For(){$t = 1, \dots, T$}{
% 			Compute $\forall k, U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$\;
% 			Transmit in channel $C(t) = \arg\max_{1\leq k \leq K} U_k(t)$\;
% 			Reward $r_{C(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
% 		}
% 		\caption{The \UCB{} algorithm for channel selection (base building block).}
% 		\label{algo:43:UCB}
% 		% \end{small} % XXX remove if needed
% 	\end{algorithm}
% 	\end{framed}
% \end{figure}
% % \end{small} % XXX remove if needed

% ----------------------------------------------------------------------
\subsection{The first heuristics: \UCB{} unaware of retransmissions}
\label{sub:43:heuristics}
% ----------------------------------------------------------------------

% \paragraph{First stage \UCB}\label{sub:43:firstStageUCB}

This first heuristic we propose is unaware of retransmission: the same channel is used for retransmissions.
The \UCB{} algorithm is implemented independently by each device, and we present it in Algorithm~\ref{algo:43:UCB}.
Note that a device using this first approach is only able to select a channel for the first transmission (using \UCB), and it uses the same channel for all the corresponding retransmissions of a packet (if retransmissions happen).

More formally, for one device, let $N_k(t)$ be the number of times the channel $k$ (for $k\in \llbracket 1, K \rrbracket$) was selected up-to time $t-1$, for $t\geq 0$
for any $t\in\mathbb{N}$,
$N_k(t) = \sum_{\tau=0}^{t-1} \mathbbm{1}(C(\tau) = k)$.
The empirical mean estimator $\widehat{\mu_k}(t)$ of channel $k$ is defined as the mean reward obtained up-to time $t-1$,
$\widehat{\mu_k}(t) = \frac{1}{N_k(t)} \sum_{\tau=0}^{t-1} r_k(\tau) \mathbbm{1}(C(\tau) = k)$,
where $r_{k}(t)$ is the reward obtained after transmission in channel $k$ at time $t$.
The upper confidence bound in each channel $k$ is defined as
$U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$.
Finally, the transmission channel at time step $t$
is the one maximizing this \UCB{} index $U_k(t)$,
as it is the one expected to be the best one at the current time step $t$,
$C(t) = \arg\max_{1\leq k \leq K} U_k(t)$.

% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Compute $\forall k, U_k(t) = \widehat{\mu_k}(t) + \sqrt{\alpha \log(t) / N_k(t)}$\;
				Transmit in channel $C(t) = \arg\max_{1\leq k \leq K} U_k(t)$\;
				Reward $r_{C(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
			\Else(\tcp*[f]{Retransmit in same channel}){
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Transmit in channel $C(t) = j$\;
			}
		}
		\caption{Heuristic: first-stage \UCB{} and retransmission in same channel.}
		\label{algo:43:UCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

\paragraph{A warning about notations: local (device) time vs global (gateway) time.}
%
In all the algorithms and notations presented in this Section, the time steps $t$ do not refer to the global time steps (as seen from the gateway), but rather the current number of uplink transmissions (or retransmissions) that was carried out by the device.
In other words, each device wakes up whenever it has to send some data, following its random emission process (here, we propose a Bernoulli process of small probability, \eg, $p=10^{-3}$), and when it woke up, it sends its packet, and will send again for at most $M$ times (\eg, $M=10$) until it receives an \Ack.
To be clear, once again, we repeat that the time steps $t$ in the following algorithms denote the number of time the device sent an uplink packet, and these uplink (re)transmissions can happen at different (real-world) times, more or less spread out in time, it does not matter to the learning part of this model and work.

@@

A device that implements the UCB algorithm is led to focus its transmissions and retransmissions in the channel which is currently identified as the best.
As explained above in Section~\ref{sub:43:motivations}, focusing in one channel could increase the collision probability in retransmissions.
We describe here the proposed heuristics for the channel selection in a retransmission. It is carried out taking into account that a device can incorporate a different channel selection strategy while being in a back-off state.
Hence, a natural question is to evaluate whether using this additional contextual information can improve the performance of a learning policy.

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
    \begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(){$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^r$}){
				Compute $\forall k, U^r_k(t) = \widehat{\mu^r_k}(t) + \sqrt{\alpha \log(t) / N_k^r(t)}$\;
				Transmit in channel $C^r(t) = \arg\max_{1\leq k \leq K} U^r_k(t)$\;
				Reward $r^r_{C^r(t)}(t) = 1$, if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: \UCB{} for retransmission.}    % A naive
		\label{algo:43:TwoUCB}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		% \KwData{$\forall k,j\in[\![1;K]\!]$, $N_k^j(t)$, $\widehat{\mu_k}^j(t)$, $B_k^j(t)$ and $U_k^j(t)$}
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\Else(\tcp*[f]{Packet retransmission with $\mathrm{UCB}^j$}){ % Retr. after trying channel $j$
				j $\leftarrow$ last channel selected by first-stage \UCB\;
				Compute $\forall k, U_k^j(t) = \widehat{\mu_k}^j(t) + \sqrt{\alpha \log(t) / N_k^j(t)}$\;
				Transmit in channel $C^j(t) = \arg\max_{1\leq k \leq K} U^j_k(t)$\;
				Reward $r^j_{C^j(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: $K$ different {\UCB}s for retransmission.}
		\label{algo:43:UCBthenKp1}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

@@

The idea behind this delay is to allow the first stage \UCB{} to start learning the best channel, before starting the second stage \UCB{} (see details in Algorithm~\ref{algo:43:UCBwithDelay}).
The number of transmissions to wait before applying the second algorithm is denoted by $\Delta$, it has to be fixed before-hand\footnote{~Choosing the value of $\Delta$ could be done by extensive benchmarks but such approach goes against the reinforcement learning idea: an heuristic should work against any problem, without the need to simulate the problem before-hand to find a good value of some internal parameter. As such, we only consider a delay of $\Delta=100$.}.
%
Note that, we use the subscript $({}^d)$ to denote the variables
related to the delayed second-stage \UCB{} algorithm.

\vspace*{-3pt}
% \begin{small} % XXX remove if needed
\begin{figure}[h!]
	\centering
	\begin{framed}
	\begin{algorithm}[H]
		% \begin{small} % XXX remove if needed
		\For(\tcp*[f]{At every time step})
		{$t = 1, \dots, T$}{
			\uIf{First packet transmission}{
				Use first-stage \UCB. %as in Algorithm~\ref{algo:43:UCB}.
			}
			\uElseIf(\tcp*[f]{Random retransmission}){$t \leq \Delta$}{
				Transmit randomly in a channel $C(t) \sim \mathcal{U}(1,\ldots,K)$.
			}
			\Else(\tcp*[f]{Packet retransmission with delayed $\mathrm{UCB}^d$}){
				Compute $\forall k, U^d_k(t) = \widehat{\mu^d_k}(t) + \sqrt{\alpha \log(t) / N_k^d(t)}$\;
				Transmit in channel $C^d(t) = \arg\max_{1\leq k \leq K} U^d_k(t)$\;
				Reward $r^d_{C^d(t)}(t) = 1$ if \emph{Ack} is received, else $0$\;
			}
		}
		\caption{Heuristic: Delayed \UCB{} for retransmission.}
		\label{algo:43:UCBwithDelay}
		% \end{small} % XXX remove if needed
	\end{algorithm}
	\end{framed}
\end{figure}
% \end{small} % XXX remove if needed

\paragraph{Other ideas that we did not explore}
%
Instead of considering a first-stage \UCB{} followed by a random channel retransmission, we could have considered a random first-stage channel retransmission followed by a second-stage \UCB.
This was not considered in our study \cite{Bonnefoi2019WCNC}, and we preferred to not add it to the experiments presented below, for three reasons: to avoid clutter in the plots, to win some time as these experiments had to run for quite a long time, but mainly because the first-stage channel selection is most important one as we illustrate by the large difference in terms of performance between the uniform channel selection and the ``only UCB'' heuristic, below in Figures~\ref{fig:43:mainExperiment1} and \ref{fig:43:mainExperiment2}.

Similarly, Philippe Mary asked me at the MoTION workshop where I presented \cite{Bonnefoi2019WCNC}, why didn't we rather used Bayes-UCB instead of \UCB{} (from \cite{Kaufmann12BUCB}), as it is known to empirically outperform \UCB{} (as illustrated for instance in \cite{darak2016bayesian,kumar2016two}).
We could have chose any bandit algorithm for the ``first stage'' component used in this Section, and for simplicity and clarity we focussed on a simple but efficient one (\UCB).
We believe that the same empirical results, but more importantly, the same conclusions, could be given if we used Bayes-UCB or other bandit algorithm for the base building block in the different heuristics proposed in Section~\ref{sub:43:heuristics}.
